Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-12-22 demystify.tex
updates

However, if the student has contacted a faculty member and that faculty is interested in the student and has made this known to the admission committee, then the student is likely to be admitted, \emph{even if the faculty does not review the application or has funding} (thus the benefit of contacting faculty as discussed in\autoref{sec:contact}). 
Similarly, if the student mentioned a faculty in their SOP, adcom might ask that faculty to look at their application and if they are interested in the student.  Even if the student has a weak profile (but still passes the minimum requirement from the university), they might be admitted if a faculty is willing to take them. Adcom members, especially in the US, are very reluctant to go against the faculty's decision (e.g., if a faculty wants to admit a student, we are not going to reject them).

@@

However, in some rare cases we do talk to each other.  For example, when there are discrepancies in evaluations, in which case the adcom chair will ask reviewers to discuss the application to reach a consensus.  We also talk to each other for interesting or strong applications, e.g., how to recruit this student or who should be the adviser. 
If the student mentioned a faculty member in their SOP, we might ask that faculty if they are interested in the student.

@@

Sometimes students would go to great lengths just to get letters from well-known professors in their school (e.g., department head or dean). But as mentioned, if these professors do not know you, their letters would likely be generic and carry little value (sometimes \red{red flags}). Moreover, a top professor at your university might not be well-known to US faculty (see more details in \autoref{sec:admin-letters} and \autoref{chap:your-school}). So save the trouble and get letters from \emph{any} professors/supervisors who know you well and can write a good letter about \emph{your} research ability. It's better to have a good personalized letter about your own research ability from someone who is less well-known than a generic/weak letter from a well-known person.

@@

Note that most places only have deadlines for the applicant, but are very flexible with the letter writers.  In many cases your LOR writers \emph{are not given any deadline}. The following lists several LOR invitation emails I received from various universities in the past few years:

\begin{emphasisbox}
    \footnotesize
    \paragraph{No deadline given} Examples with no deadline given:

@@

    Your letter of recommendation is due by \textbf{12 p.m. (Noon) EST on December 10, 20XX}. Thank you for taking the time to respond.
\end{emphasisbox}

@@

\begin{commentbox}[Thanh:]
    Due to academic culture, professors in Vietnam usually aim for (international) journals instead of conferences. Could you give some tips on how to know whether a journal is good (CSrankings, unfortunately, only consider conferences)?
    \tcblower
    \textbf{Vu}: One way is looking at what well-known researchers publish. For example, if you are interested in a field X, you can use CSRankings to look at active faculty in X, and then look at their websites to see what journals they publish at.
\end{commentbox}
\paragraph{What if I don't have any pubs} It is understandable that many students do not have the opportunity to publish papers. Thus, other writings, even those under submissions or rejected, would still be good and better than nothing.  Be sure to upload your papers with your application (\autoref{sec:writing-sample}) and talk about them in your SOP (\autoref{chap:sop}).

@@

\paragraph{I am not the first author} Being the first author is good because it indicates you own the work. However, it's perfectly OK to be second or third or last.  It is difficult to publish a good paper and so being a co-author is still a good sign about your research experience. In any case (and especially in the case when you're not the first author), you should explain the work and your contribution.  Better yet, have your LoR writers talk about your contribution in their letters.

@@

Note that adcom reviewers might not be familiar with all research labs, especially those outside the US For example, while VinAI is  well-known in Vietnam and potentially in Asia, it might not be very well-known in the US. So you or your LoR writers should explicitly say something about them in your statement.  In general, if you did some good research work, then you should mention that in your SOP and ask your supervisor to write about it in their LOR (\autoref{chap:lor}).

@@

\section{Competitions} Winning \emph{internationally recognized competitions} can demonstrate your research potential.
For example, participating in Math Olympiads if you want to do theory or winning ACM programming contests if you want to ``build'' systems, e.g., software analysis.  
So do talk about them in your SOP (\autoref{chap:sop}) and have your LoR writers mention them in their letters (\autoref{chap:lor}).

@@

\section{Interviews}

@@

The timeline for interviews varies.  Faculty set up interviews based on their (busy and erratic) schedule. Do not be surprised if you get an interview invitation at the last minute. Some schools do not do interviews at all (\autoref{sec:no-interview}).

@@

@@

\section{Contacting a prof. How to get a desired reply?}\label{sec:contact}

@@

    \item \textbf{Don't CALL}. Not related to email but sometimes students get desperate and call the prof.  This is a big no-no, especially for CS ppl who often prefer email over phone calls. Seriously though, don't call any prof. unless they ask you to. 
\end{itemize}

@@

\item \textbf{I am not taking students but I co-advise/can serve on your committee} While this might sound good, it's generic because it says once you're admitted and have an adviser, then contact me again.

@@

In the schools I've been at, the application deadline is in December, and adcom meets when school starts in mid or end of January. Adcom chair sends out review assignments to adcom members, about 30ish per faculty (\autoref{sec:adcom}). We usually have about 2 weeks to review all applications. As mentioned in~\autoref{sec:adcom-discuss}, adcom members review applications individually and independently. We only discuss when there are disagreements (adcom chair will determine which applications to discuss).

I typically reserve a whole day to review all applications.
On average, I spend about 10--15 minutes reviewing each application (less for clear rejections and more for potential acceptances). While this seems short, it is not that difficult to tell if an application is good or bad.  In fact, this
is twice what other faculty spend on average, e.g., Philip Guo spent \href{https://pg.ucsd.edu/PhD-application-tips.htm}{3--5 minutes} per application.

@@

Finally, I enter my decision, which is usually either a rejection or an offer of admission with full funding (e.g., from TA). I also recommend very strong candidates for the University Presidential fellowship., which is a fellowship from GMU that provides funding similar to an RA (\autoref{sec:funding}).
Note that while the system has other decision options, eg., admit without funding, provisional admission (e.g., if they need to take some courses), I rarely use them, simply because we want to fully fund our PhD students.

@@

\begin{commentbox}[DK:]
  I was amused to read this as if I recall correctly, you never called me by my first name when you were at UNM. You always called me Prof. And, many times, I would jokingly call you back as Prof. Vu.
  \tcblower
  \textbf{Vu}: Yes, for some reason I enjoy calling you ``Professor'' (without appending a last or first name).  The use of Prof. Vu may have foreshadowed my future in academia.
\end{commentbox}

@@

Do not include ranking (e.g., Assistant, Associate, Scientist, ...) when referring to someone. I see many international students include a lengthy title of people they know, e.g., \emph{I am advised by Asst. Prof. X, and I also collaborate with Distinguished Scientist Y}.  This is \emph{not necessary} and makes it look like you're trying to show off your connections. These nuances represent some cultural and academic differences that you may encounter and will gradually adapt to. More on cultural differences in \autoref{sec:cultural}.

>>>>

However, if the student has contacted a faculty member and that faculty is interested in the student and has made this known to the admission committee, then the student is likely to be admitted, \emph{even if the faculty does not review the application or has funding} (thus the benefit of contacting faculty as discussed in~\autoref{sec:contact}).

Similarly, if the student mentioned a faculty in their SOP, adcom might ask that faculty to look at their application and if they are interested in the student.  Even if the student has a weak profile (but still passes the minimum requirement from the university), they might be admitted if a faculty is willing to take them. Adcom members, especially in the US, are very reluctant to go against the faculty's decision (e.g., if a faculty wants to admit a student, we are not going to reject them).

@@

Not really, we typically review applications independently and do not talk to each other.  This is to avoid biasing, e.g., if one reviewer says they want to accept, the other might feel pressured to accept as well.

However, when there are discrepancies in evaluations, the adcom chair will ask reviewers to discuss the application to reach a consensus.  We might also talk to each other for interesting or strong applications, e.g., how to recruit this student or who should be the adviser. 
If the student mentioned a faculty member in their SOP, we might ask that faculty if they are interested in the student.

Note that other disciplines might have different practices, e.g., they might select a top list of applicants and then discuss them in a meeting to determine who to interview. In CS, both the reviews and interviews are often done independently (\autoref{sec:interviews}).

\begin{commentbox}[A prof. in Physics at an R1 university:]
  We have a pretty well fleshed out grading rubric for applications that has categories like grades, research, writing ability, etc. I would say our rubric is weighted about 1/2 on academics (research, LORs, grades) and 1/2 on the idea of ``grit'' or ``resilience'' (engagement, leadership, working through obstacles).\\

  The rubric helps a lot to standardize how committee members grade, and speeds things up a bit because you know what to look for. We spent what seemed like forever on the details of the grading system (e.g., what does a a score of `3' vs a `2' in writing look like?) but now it's very helpful.\\

  We also do roughly three rounds of selection: a first ``triage'' round to determine the top ~100 applications, a second round to determine about 25 people to interview, and then a third round to decide the actual offers. That also helps to speed things up a bit, since in the first round with all the applications you can move fairly quickly since you just need to sort into ``good'' and ``bad''. By the time we're getting into the details and reading everything more closely in rounds 2 and 3 most of the applications have been removed from consideration. So for this method I do maybe 5 minutes per app in round 1, but closer to 20 minutes per app in round 2, and usually round 3 is long discussions about specific people.   
\end{commentbox}

@@

Sometimes students would go to great lengths just to get letters from ``top'' professors in their school (e.g., department head or dean). But as mentioned, if these professors do not know you, their letters would likely be generic and carry little value (sometimes \red{red flags}). Moreover, a top professor at your university might not be well-known to US faculty (see more details in \autoref{sec:admin-letters} and \autoref{chap:your-school}). So save the trouble and get letters from \emph{any} professors/supervisors who know you well and can write a good letter about \emph{your} research ability. It's better to have a good personalized letter about your own research ability from someone who is less well-known than a generic/weak letter from a well-known person.

@@

Note that most places only have deadlines for the applicant, but are very flexible with the letter writers.  In many cases your LOR writers \emph{are not given any deadline}. \autoref{fig:fig:lor-invitations} lists several LOR invitation emails I received from various universities in the past few years.

\begin{figure}
\begin{emphasisbox}
    \footnotesize
    \paragraph{No deadline given} Examples with no deadline given:

@@

    Your letter of recommendation is due by \textbf{12 p.m. (Noon) EST on December 10, 20XX}. Thank you for taking the time to respond.
\end{emphasisbox}
\caption{Examples of LOR invitation emails}\label{fig:lor-invitations}
\end{figure}

@@

\begin{commentbox}[Thanh:]
    Due to academic culture, professors in Vietnam usually aim for (international) journals instead of conferences. Could you give some tips on how to know whether a journal is good (CSrankings, unfortunately, only consider conferences)?
    \tcblower
    \textbf{Vu}: One way is looking at what well-known researchers publish. For example, if you are interested in a field X, you can use CSRankings to look at active faculty in X, and then look at their websites to see what journals they publish at.
\end{commentbox}

\paragraph{What if I don't have any pubs} It is understandable that many students do not have the opportunity to publish papers. Thus, other writings, even those under submissions or rejected, would still be good and better than nothing.  Be sure to upload your these with your application (\autoref{sec:writing-sample}) and talk about them in your SOP (\autoref{chap:sop}).

@@

\paragraph{I am not the first author} Being the first author is good because it indicates you own the work. However, it's \emph{perfectly OK to be second or third or even last}.  Adcom members know it is difficult to publish a good paper, and so being a co-author is still a good sign about your research experience. In any case, especially in the case when you're not the first author, you should explain the work and your contribution.  Better yet, have your LoR writers (\autoref{sec:help-your-LOR-writers}) talk about your work and contribution in their letters.

@@

Note that adcom reviewers might not be familiar with all research labs, especially those outside the US. For example, while VinAI is  well-known in Vietnam and potentially in Asia, it might not be very well-known in the US. So you or your LoR writers should explicitly say something about them in your statement.  In general, if you did some good research work, then you should mention that in your SOP and ask your supervisor to write about it in their LOR (\autoref{chap:lor}).

@@

\section{Competitions} Winning \emph{internationally recognized competitions} can demonstrate your research potential.
For example, participating in Math Olympiads if you want to do theory or winning ACM programming contests if you want to ``build'' systems, e.g., software analysis.  
So do talk about them in your SOP (\autoref{chap:sop}) and have your LoR writers mention them in their letters (\autoref{sec:help-your-LOR-writers}).

@@

\section{Interviews}\label{sec:interviews}

@@

The timeline for interviews varies.  Faculty set up interviews based on their (busy and erratic) schedule. Some try to get interviews done before the holidays, while others do them after the holidays.
Do not be surprised if you get an interview invitation at the last minute. Some profs. are informal and may just email you to chat (e.g., \emph{``could you chat in an hr?''}), while others might schedule a formal interview (e.g., \emph{``can you chat at 2 pm on Friday or 10 am on Monday?''}).

Some schools do not do interviews at all (\autoref{sec:no-interview}). They review applications and make decisions based on them.  If you do not get an interview, it does not mean you're out.

@@

\myepigraphsimpsons{``It’s not easy being a mentor. But it sure is funny to watch other people screw up.''}

@@

\section{Contacting a Prof. How to get a desired reply?}\label{sec:contact}

@@

    \item \textbf{Do Not Call}. Not related to email but sometimes students get desperate and call the prof.  This is a big no-no, especially for CS ppl who often prefer email over phone calls. 
\end{itemize}

@@

\item \textbf{I am not taking students but I co-advise/can serve on your committee.} While this might sound good, it's still generic because it says \emph{once you're admitted and have an adviser}, then contact me again.

@@

In the schools I've been at, the application deadline is in Dec, and adcom meets when school starts in mid or end of Jan. Adcom chair sends out review assignments to adcom members, about 30ish per faculty (\autoref{sec:adcom}). We usually have about 2 weeks to review all applications. As mentioned in~\autoref{sec:adcom-discuss}, adcom members review applications individually and independently. We only discuss when there are disagreements (adcom chair will determine which applications to discuss).

I typically reserve a whole day (or two days) to review all applications.
On average, I spend about 10--15 minutes reviewing each application (less for clear rejections and more for potential acceptances). While this seems short, it is not that difficult to tell if an application is good or bad.  In fact, this
is twice what other faculty spend on average, e.g., Philip Guo spent \href{https://pg.ucsd.edu/PhD-application-tips.htm}{3--5 minutes} per application.

@@

Finally, I enter my decision, which is usually either a rejection or an offer of admission with full funding (e.g., from TA). I also recommend very strong candidates for the University Presidential fellowship., which is a fellowship from GMU that provides funding similar to an RA (\autoref{sec:funding}).
Note that while the system has other decision options, eg., admit without funding, provisional admission (e.g., if they need to take some courses), I do not use them, simply because we either reject or admit with funding.

Note that I \emph{do not} need to interview a student to make a decision.  I can tell from the application if they are strong or not. However, if I want to recruit a student, I will ask them to chat with me. 
This is quite different than in other discipline where reviews, interviews, and selection are done in multiple rounds (see~\autoref{sec:adcom-discuss} for an example in Physics).

Of course my recommendation is just one of the three or four faculty who review the application.  The adcom chair will compile all recommendations and make a decision based on them (\autoref{sec:adcom-discuss}).
If there are disagreements, adcom chair will ask the reviewers to discuss the application. 
Unfortunately, even if all revierwers recommend a student, they might not be admitted (\autoref{sec:why-rejected}) if there are too many students or they are not a good fit for the program (e.g., no one is willing to advise them).

@@

\begin{commentbox}[Kapur:]
  I was amused to read this as if I recall correctly, you never called me by my first name when you were at UNM. You always called me Prof. And, many times, I would jokingly call you back as Prof. Vu.
  \tcblower
  \textbf{Vu}: Yes, for some reason I enjoy calling you ``Professor'' (without appending a last or first name).  The use of Prof. Vu may have foreshadowed my future in academia.
\end{commentbox}

@@

Do not include ranking (\autoref{sec:faculty-types}), e.g., Assistant, Associate, Scientist, ..., when referring to someone. I see many international students include a lengthy title of people they know, e.g., \emph{I am advised by Asst. Prof. X, and I also collaborate with Distinguished Scientist Y}.

This is \emph{not necessary} and makes it look like you're trying to show off your connections. These nuances represent some cultural and academic differences that you may encounter and will gradually adapt to. More on these differences in \autoref{sec:cultural}.
